---
title: "COVID19 Time Series Analysis, Worldwide and U.S."
author: "Andy Hasselwander"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 2
---

Source data: 2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by Johns Hopkins CSSE; https://github.com/CSSEGISandData/COVID-19

Source code: https://github.com/opencedar/covid19

The visualizations in this document are heavily indebted to Edward Tufte and his use of sparklines—small, clutter-free time series lines—to show how many different panels or categories of data are changing through time; check out https://www.edwardtufte.com/bboard/q-and-a-fetch-msg?msg_id=0001OR. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

####HOPKINS DATABASE####

# Get deaths, confirmed, recovered by country, region by date. These are three separate databases 

#------------------------------------------------------------------------------------------------
rm(list = ls())
library(tidyverse)
library(zoo)
library(scales)
source("functions.R")

daily_vector <- seq(from = as.Date("2020-01-22"), (lubridate::today()-1), "day")
daily_vector <- format(daily_vector, "%m-%d-%Y")

for(i in 1:length(daily_vector)) {
  day_get <- daily_vector[i]
  if(i==1) {
    hopkins_df <- read.csv(paste0("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/", daily_vector[i],".csv"))
    hopkins_df <- data.frame(date = rep(day_get, times = nrow(hopkins_df)), hopkins_df) 
    colnames(hopkins_df) <- gsub("\\.", "_", colnames(hopkins_df))
  } else {
    df <- read.csv(paste0("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/", daily_vector[i],".csv"))
    df <- data.frame(date = rep(day_get, times = nrow(df)), df) 
    colnames(df) <- gsub("\\.", "_", colnames(df))
    hopkins_df <- bind_rows(hopkins_df, df)
  }
}
hopkins_df$date <- lubridate::mdy(hopkins_df$date)
hopkins_df$Last_Update <- lubridate::mdy_hms(hopkins_df$Last_Update)



state_key_o <- read.csv("https://raw.githubusercontent.com/jasonong/List-of-US-States/master/states.csv", stringsAsFactors = FALSE)
state <- data.frame(State = state_key_o$State, Abbreviation = state_key_o$State)
state_key <- rbind(state_key_o, state)

x <- hopkins_df

#Fix China, Iran
x$Country_Region[x$Country_Region == "Mainland China"] <- "China"
x$Country_Region[x$Country_Region == "Iran (Islamic Republic of)"] <- "Iran"
x$Country_Region[x$Country_Region == "Republic of Korea"] <- "Korea, South"
x$Country_Region[x$Country_Region == "South Korea"] <- "Korea, South"


x$state_from_county <- str_extract(x$Province_State, '(A[LKSZR])|(C[AOT])|(D[EC])|(F[ML])|(G[AU])|(HI)|(I[DLNA])|(K[SY])|(LA)|(M[EHDAINSOT])|(N[EVHJMYCD])|(MP)|(O[HKR])|(P[WAR])|(RI)|(S[CD])|(T[NX])|(UT)|(V[TIA])|(W[AVIY])')
x <- merge(x, state_key, all.x=TRUE, all.y = FALSE, by.x = 'state_from_county', by.y = "Abbreviation")
x$Province_State[!is.na(x$State)] <- x$State[!is.na(x$State)] 

x <- x %>% select(colnames(hopkins_df))


covid_ets <- x %>% group_by( Country_Region, Province_State, date) %>% summarise(confirmed = sum(Confirmed, na.rm=TRUE), deaths = sum(Deaths, na.rm=TRUE), recovered = sum(Recovered, na.rm = TRUE), active = sum(Active))

write.csv(covid_ets, "covid_ets.csv")

#covid_country_ets data set is the primary country-level time series data set
covid_country_ets <- covid_ets  %>% 
  group_by(Country_Region, date) %>% 
  summarise(confirmed = sum(confirmed, na.rm=TRUE), deaths = sum(deaths, na.rm=TRUE), recovered = sum(recovered, na.rm = TRUE)) %>% 
  mutate(
    #create daily percent growth variables
    confirmed_growth_rate = c(NA,diff(confirmed))/lag(confirmed, 1),
    death_growth_rate = c(NA,diff(deaths))/lag(deaths, 1),
    death_rate = deaths / confirmed,
    active = confirmed - deaths,
    active_lag14 = dplyr::lag(confirmed - deaths, 14, 0),
    active_lag7 = dplyr::lag(confirmed - deaths, 7, 0),
    confirmed_lag7 = dplyr::lag(confirmed, 7, 0),
    confirmed_lag14 = dplyr::lag(confirmed, 14, 0),
    new_confirmed = c(NA, diff(confirmed)),
    new_deaths = c(NA, diff(deaths)),
    new_confirmed_lag7 = dplyr::lag(new_confirmed,7,0),
    new_confirmed_lag14 = dplyr::lag(new_confirmed,14,0),
    new_confirmed_ma7 = rollmean(x=new_confirmed, 7, fill=NA, align = "right"),
    new_confirmed_ma14 = rollmean(x=new_confirmed, 14, fill=NA, align="right")
  )

#replace NaNs w 0s
covid_country_ets$confirmed_growth_rate[is.nan(covid_country_ets$confirmed_growth_rate)] <- 0
covid_country_ets$death_growth_rate[is.nan(covid_country_ets$death_growth_rate)] <- 0

#create country 
covid_country_ets <- covid_country_ets %>% mutate(
  conf_growth_rate_ma=rollapply(confirmed_growth_rate,5,mean,align='right',fill=NA),
  death_growth_rate_ma=rollapply(death_growth_rate,5,mean,align='right',fill=NA)) %>%
  ungroup()

#Filter out growth w < 100 base
covid_country_ets$conf_growth_rate_ma[covid_country_ets$confirmed <100] <- NA
covid_country_ets$death_growth_rate_ma[covid_country_ets$deaths <100] <- NA
covid_country_ets$death_rate[covid_country_ets$deaths <100] <- NA

#make rows with confimed cases and NA new_confirmed have 0 not NA new_confirmed
covid_country_ets$new_confirmed[is.na(covid_country_ets$new_confirmed) & !is.na(covid_country_ets$confirmed)] <- 0


#Only get countries with >1000 cases today
current_cases <- covid_country_ets %>% group_by(Country_Region) %>% summarise(current_cases = max(confirmed, na.rm=TRUE))
five_hundred <- current_cases$Country_Region[current_cases$current_cases >= 1000]

covid_country_ets <- covid_country_ets %>% filter(Country_Region %in% five_hundred)

#Create a panel data set
library(plm)
covid_plm <- pdata.frame(covid_country_ets, index=c("Country_Region", "date"))
#create diff variables
covid_plm$diff_confirmed <- diff(covid_plm$confirmed, 1)
covid_plm$diff_deaths <- diff(covid_plm$deaths, 1)


#######
```
# Worldwide

## Worldwide Summary

Sorted by total number of cases. Percent growth in total cases in the past seven days is last column.
```{r echo=FALSE, message=FALSE, warning=FALSE}

covid_country_fc <- data.frame(covid_country_ets)
#Only get country-days with >= 100 cases
covid_country_fc <- covid_country_fc %>% filter(confirmed >= 100)
covid_country_fc <- covid_country_fc %>% group_by(Country_Region) %>% mutate(days_since_100 = dplyr::row_number())

# Create NA versions of variables
covid_country_fc$deaths_na <- covid_country_fc$deaths
covid_country_fc$deaths_na[covid_country_fc$deaths_na <= 0] <- NA
covid_country_fc$confirmed_na <- covid_country_fc$confirmed
covid_country_fc$confirmed_na[covid_country_fc$confirmed_na <= 0] <- NA
covid_country_fc$days_since_100_na <- covid_country_fc$days_since_100
covid_country_fc$days_since_100_na[covid_country_fc$days_since_100_na <= 0] <- NA
covid_country_fc$confirmed_lag7_na <- covid_country_fc$confirmed_lag7
covid_country_fc$confirmed_lag7_na[covid_country_fc$confirmed_lag7_na <= 0] <- NA
covid_country_fc$confirmed_lag14_na <- covid_country_fc$confirmed_lag14
covid_country_fc$confirmed_lag14_na[covid_country_fc$confirmed_lag14_na <= 0] <- NA
covid_country_fc$new_confirmed_lag7_na <- covid_country_fc$new_confirmed_lag7
covid_country_fc$new_confirmed_lag7_na[covid_country_fc$new_confirmed_lag7_na <= 0] <- NA
covid_country_fc$new_confirmed_lag14_na <- covid_country_fc$new_confirmed_lag14
covid_country_fc$new_confirmed_lag14_na[covid_country_fc$new_confirmed_lag14_na <= 0] <- NA
covid_country_fc$new_confirmed_na <- covid_country_fc$new_confirmed
covid_country_fc$new_confirmed_na[covid_country_fc$new_confirmed_na <= 0] <- NA
covid_country_fc$new_confirmed_ma7_na <- covid_country_fc$new_confirmed_ma7
covid_country_fc$new_confirmed_ma7_na[covid_country_fc$new_confirmed_ma7_na <= 0] <- NA
covid_country_fc$new_confirmed_ma14_na <- covid_country_fc$new_confirmed_ma14
covid_country_fc$new_confirmed_ma14_na[covid_country_fc$new_confirmed_ma14_na <= 0] <- NA

#create country summary
country_summary <- covid_country_fc %>%
  group_by(Country_Region) %>% 
  slice(which.max(days_since_100)) %>% 
  select(Country_Region, days_100 = days_since_100, conf = confirmed, deaths, new_conf=new_confirmed, conf_lag7=confirmed_lag7) %>%
  arrange(desc(conf))

#create worldwide totals
world <- data.frame(Country_Region = "World", 
                    days_100 = mean(country_summary$days_100),
                    conf = sum(country_summary$conf),
                    deaths = sum(country_summary$deaths),
                    new_conf = sum(country_summary$new_conf),
                    conf_lag7 = sum(country_summary$conf_lag7))

country_summary <- bind_rows(country_summary, world)   

country_summary <- country_summary %>% mutate(l7_rate = (conf / conf_lag7)-1)

country_summary$l7_rate <- scales::percent(country_summary$l7_rate, accuracy = 4, big.mark = ",")


knitr::kable(country_summary, caption = "Worldwide Summary", digits  = 0, align = c("l",rep("r", times = 7)))

```

## Ln (Seven-Day-Moving-Average New Cases) Impact on Ln (New Cases)

In other words, elasticity. How does this elasticity change through time, from days since the 100th case?

An elasticity under 1 indicates that over a seven-day period, new cases are decreasing.

The black line shows the best curve fit for elasticity changing over time. All countries generally are moving to cap the rate of exponential growth. Countries above the line are doing worse than average, and those below the line are doing better than average. A rate below 1 indicates that new cases are declining over an average 7-day period.

```{r echo = FALSE, message=FALSE, warning=FALSE}
library(broom)
#Remove NAs

confirmed_models <- covid_country_fc %>% group_by(Country_Region) %>% do(model = lm(log(new_confirmed_na) ~  log(new_confirmed_ma7_na) -1, data = .))
confirmed_tidy_model <- confirmed_models %>% tidy(model)

days_since_100_by_country <- covid_country_fc %>% group_by(Country_Region) %>% summarize(days_since_100 = max(days_since_100))

seven_day_elasticity_by_country <- data.frame(country = confirmed_tidy_model$Country_Region, days_since_100  = as.numeric(days_since_100_by_country$days_since_100), lag_7_elasticity = confirmed_tidy_model$estimate)


cases_7_day_model_ww <- lm(lag_7_elasticity ~ log(days_since_100),data = seven_day_elasticity_by_country)

seven_day_elasticity_by_country$prediction <-  cases_7_day_model_ww$coefficients[[1]] + cases_7_day_model_ww$coefficients[[2]] * log(seven_day_elasticity_by_country$days_since_100)

seven_day_elasticity_by_country_no_outliers <- seven_day_elasticity_by_country %>% filter(lag_7_elasticity <2)

require(ggrepel)
ggplot(seven_day_elasticity_by_country_no_outliers, aes(x=days_since_100)) +
  geom_point(shape=1, show.legend = FALSE, (aes(y=lag_7_elasticity, color = country))) +    # Use hollow circles
  geom_line(aes( y = prediction)) +
  geom_label_repel(data=  seven_day_elasticity_by_country_no_outliers, aes(x=days_since_100, y=lag_7_elasticity, label = country), size=2) + 
  theme(legend.position = "none") 

knitr::kable(seven_day_elasticity_by_country_no_outliers %>% mutate(residual = lag_7_elasticity - prediction) %>% arrange(days_since_100, residual), caption = "Countries by Predicted vs. Actual Lag 7 New Case Elasticity on Today's Cases", digits = 2)

actuals <- covid_country_fc %>% select(country = Country_Region, date, days_since_100 = days_since_100, confirmed, new_confirmed, new_confirmed_lag7)

prediction <- expand.grid(country = unique(seven_day_elasticity_by_country$country), days_since_100 = 1:100) %>% 
  arrange(country, days_since_100)

prediction$prediction <- predict.lm(cases_7_day_model_ww, prediction)

prediction <- merge(prediction, actuals, on.x = c("country", "days_since_100"), all.x = TRUE, all.y = FALSE)


```

### Comparisons with averages

```{r echo = FALSE, message=FALSE, warning=FALSE}
country_residuals <- seven_day_elasticity_by_country_no_outliers %>% mutate(ww_residual = lag_7_elasticity - prediction)  %>% arrange(days_since_100, ww_residual)

knitr::kable(country_residuals, caption = "Countries by Predicted vs. Actual Lag 7 New Case Elasticity on Today's Cases", digits  = 2)

```


## Forecast New Cases by Country

We estimate new cases by date, to see when countries will peak, based on the worldwide curve fit.

```{r echo = FALSE, message=FALSE, warning=FALSE}

non_na_countries <- covid_country_ets %>% group_by(country = Country_Region) %>% summarise(nonna_count = sum(!is.na(confirmed)))
non_na_countries <- non_na_countries %>% filter(nonna_count >= 20)

#Get country populations (2016 is latest data I can find easily)
country_pop <- read.csv("https://raw.githubusercontent.com/datasets/population/master/data/population.csv", stringsAsFactors = FALSE)
country_pop <- country_pop %>%  
  filter(Year == 2016) %>%
  select(country = Country.Name, population = Value)

country_pop$country[country_pop$country == "United States"] <- "US"

actuals <- covid_country_fc %>% 
  filter(Country_Region %in% non_na_countries$country) %>%
  select(country = Country_Region, date, days_since_100 = days_since_100, confirmed, new_confirmed, new_confirmed_ma7) 

prediction_ww <- expand.grid(country = unique(seven_day_elasticity_by_country$country), days_since_100 = 1:250) %>% 
  arrange(country, days_since_100) %>%
  filter(country %in% non_na_countries$country) %>%
  inner_join( country_pop, by = c("country" = "country")) 

prediction_ww$ww_coefficient <- cases_7_day_model_ww$coefficients[[1]] + cases_7_day_model_ww$coefficients[[2]] *                    log(prediction_ww$days_since_100)

country_residuals <- country_residuals %>% select(country, ww_residual)

prediction_ww <-  left_join(prediction_ww, country_residuals, by = c('country' = 'country')) 

# Try new model

covid_country_fc_p <- covid_country_fc %>% 
  group_by(Country_Region) %>% 
  mutate(lag7_elasticity = (log(new_confirmed_na)/ log(new_confirmed_lag7_na))) %>%
  filter(is.finite(lag7_elasticity))

cases_7_day_model_ww_2 <- lm(lag7_elasticity ~ log(days_since_100_na),data = covid_country_fc_p)

prediction_ww$new_ww_coefficient <- cases_7_day_model_ww_2$coefficients[[1]] + cases_7_day_model_ww_2$coefficients[[2]] *                    log(prediction_ww$days_since_100)

prediction_ww <- prediction_ww %>% mutate(ww_coefficient_adj = ww_coefficient + ww_residual, new_ww_coefficient_adj = new_ww_coefficient + ww_residual)

prediction_ww <- dplyr::left_join(prediction_ww, actuals, by = c('country', 'days_since_100')) 

sd_effect <- .03

# loop through df

for(i in 1:length(unique(prediction_ww$country))) {
  predict <- prediction_ww %>% filter(country == unique(prediction_ww$country)[i])
  predict$forecast_new <- round(forecast_forward(predict$new_confirmed, (predict$ww_coefficient_adj - sd_effect), 7, 
                                                 predict$population), digits = 0)
  if(i==1) {predict_df_ww <- predict} else {predict_df_ww <- rbind(predict_df_ww, predict)}
}

for(j in 1:length(unique(predict_df_ww$country))) {
  df_j <- predict_df_ww %>% filter(country == unique(predict_df_ww$country)[j])
  date_j <- date_forward(df_j$date)
  if(j==1) {complete_date <- date_j} else {complete_date <- c(complete_date, date_j)}
}

predict_df_ww$date <- complete_date

forecast_summary <- predict_df_ww %>% group_by(country) %>% 
  summarise(total_cases = sum(forecast_new), peak_new_cases = max(forecast_new)) 

max_date <- predict_df_ww %>% group_by(country) %>%
    slice(which.max(forecast_new)) %>% select(date)

forecast_summary <- merge(forecast_summary, max_date, on = "country")

forecast_summary <- forecast_summary %>% arrange(date)

forecast_summary <- forecast_summary %>% inner_join(country_pop, by = c("country" = "country"))  %>% mutate(perc_pop_infected = total_cases / population)

forecast_summary$perc_pop_infected <- scales::percent(forecast_summary$perc_pop_infected, accuracy = .1)
forecast_summary$total_cases <- scales::comma(forecast_summary$total_cases)
forecast_summary$peak_new_cases <- scales::comma(forecast_summary$peak_new_cases)
forecast_summary$population <- scales::comma(forecast_summary$population)

knitr::kable(forecast_summary, caption= "Forecast Peak New Cases by Country", align = c("l","r","r","r","r","r"), padding = 1
             )


```

## Forecast New Cases WW Total

```{r echo = FALSE, message=FALSE, warning=FALSE}

forecast_ww <- group_by(predict_df_ww, date) %>%
  summarise(forecast_new = sum(forecast_new)) %>%
  mutate(total_cases =  cumsum(forecast_new))

ggplot(forecast_ww[1:150,], aes(x=date, y=total_cases)) +
  geom_line() +
  ggtitle(paste0("WW Confirmed COVID19 Case Forecast as of ", format(Sys.time(), '%d %B %Y'))) + 
  scale_y_continuous(label=comma)

max_new <- forecast_ww[which.max(forecast_ww$forecast_new),]
max_new$forecast_new <- comma(max_new$forecast_new)
max_new$total_cases <- comma(max_new$total_cases)

knitr::kable(max_new, caption = "Peak Daily New Cases Worldwide and Total on That Day")

ggplot(forecast_ww[1:150,], aes(x=date, y=forecast_new)) +
  geom_line() +
  ggtitle(paste0("Worldwide New COVID19 Case Forecast as of ", format(Sys.time(), '%d %B %Y'))) + 
  scale_y_continuous(label=comma)


```


## Sparklines

### Confirmed Cases

```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=confirmed, group = Country_Region)) +
  geom_line() +
  facet_wrap(~ Country_Region) +
  ggtitle(paste0("Confirmed COVID19 Cases Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

### Deaths

```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=deaths, group = Country_Region)) +
  geom_line() +
  facet_wrap(~ Country_Region) +
  ggtitle(paste0("Cumulative COVID19 Deaths Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```


### Confirmed Growth Rate 5-Day Moving Average

```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=conf_growth_rate_ma, group = Country_Region)) +
  geom_line() +
  facet_wrap(~ Country_Region) +
  ggtitle(paste0("5-Day MA Confirmed Growth Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

### Death Rate 

```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=death_rate, group = Country_Region)) +
  geom_line() +
  facet_wrap(~ Country_Region) +
  ggtitle(paste0("Death Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```



