---
title: "COVID19 Time Series"
author: "Andy Hasselwander"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
---

Source data: 2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by Johns Hopkins CSSE; https://github.com/CSSEGISandData/COVID-19

Source code: https://github.com/opencedar/covid19

This visualization is heavily indebted to Edward Tufte and his use of sparklines—small, clutter-free time series lines—to show how many different panels or categories of data are changing through time; check out https://www.edwardtufte.com/bboard/q-and-a-fetch-msg?msg_id=0001OR. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(zoo)
#Get data
require(tidyverse)
####HOPKINS DATABASE####

# Get deaths, confirmed, recovered by country, region by date. These are three separate databases 

#------------------------------------------------------------------------------------------------

deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Deaths.csv", stringsAsFactors = FALSE)
deaths_thin <- deaths %>% gather("date", "deaths", 5:ncol(deaths))
deaths_thin$date <- substring(deaths_thin$date, 2)
deaths_thin$date <- lubridate::mdy(deaths_thin$date)

confirmed <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv", stringsAsFactors = FALSE)
confirmed_thin <- confirmed %>% gather("date", "confirmed", 5:ncol(confirmed))
confirmed_thin$date <- substring(confirmed_thin$date, 2)
confirmed_thin$date <- lubridate::mdy(confirmed_thin$date)

recovered <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Recovered.csv", stringsAsFactors = FALSE)
recovered_thin <- recovered %>% gather("date", "recovered", 5:ncol(recovered))
recovered_thin$date <- substring(recovered_thin$date, 2)
recovered_thin$date <- lubridate::mdy(recovered_thin$date)

covid_ets <- merge(deaths_thin, confirmed_thin, on= colnames(confirmed_thin)[1:5])
covid_ets <- merge(covid_ets, recovered_thin, on = colnames(covid)[1:5])

state_key <- read.csv("https://raw.githubusercontent.com/jasonong/List-of-US-States/master/states.csv", stringsAsFactors = FALSE)
state <- data.frame(State = state_key$State, Abbreviation = state_key$State)
state_key <- rbind(state_key, state)

x <- covid_ets
x$state_from_county <- str_extract(x$Province.State, '(A[LKSZR])|(C[AOT])|(D[EC])|(F[ML])|(G[AU])|(HI)|(I[DLNA])|(K[SY])|(LA)|(M[EHDAINSOT])|(N[EVHJMYCD])|(MP)|(O[HKR])|(P[WAR])|(RI)|(S[CD])|(T[NX])|(UT)|(V[TIA])|(W[AVIY])')
x <- merge(x, state_key, all.x=TRUE, all.y = FALSE, by.x = 'state_from_county', by.y = "Abbreviation")
x$Province.State[!is.na(x$State)] <- x$State[!is.na(x$State)] 

x <- x %>% select(colnames(covid_ets))

covid_ets <- x %>% group_by( Country.Region, Province.State, date) %>% summarise(confirmed = sum(confirmed, na.rm=TRUE), deaths = sum(deaths, na.rm=TRUE), recovered= sum(recovered, na.rm=TRUE))

write.csv(covid_ets, "covid_ets.csv")

covid_country_ets <- covid_ets  %>% 
  group_by(Country.Region, date) %>% 
  summarise(confirmed = sum(confirmed, na.rm=TRUE), deaths = sum(deaths, na.rm=TRUE), recovered= sum(recovered, na.rm=TRUE)) %>% 
  mutate(
    #create daily percent growth variables
    confirmed_growth_rate = c(NA,diff(confirmed))/lag(confirmed, 1),
    death_growth_rate = c(NA,diff(deaths))/lag(deaths, 1),
    recovered_growth_rate = c(NA,diff(recovered))/lag(recovered, 1),
    death_rate = deaths / confirmed,
    recovered_rate = recovered / confirmed
   
  )

#replace NaNs w 0s
covid_country_ets$confirmed_growth_rate[is.nan(covid_country_ets$confirmed_growth_rate)] <- 0
covid_country_ets$death_growth_rate[is.nan(covid_country_ets$death_growth_rate)] <- 0
covid_country_ets$recovered_growth_rate[is.nan(covid_country_ets$recovered_growth_rate)] <- 0

covid_country_ets <- covid_country_ets %>% mutate(
    conf_growth_rate_ma=rollapply(confirmed_growth_rate,5,mean,align='right',fill=NA),
    death_growth_rate_ma=rollapply(death_growth_rate,5,mean,align='right',fill=NA),
    recovered_growth_rate_ma=rollapply(recovered_growth_rate,5,mean,align='right',fill=NA)) %>%
ungroup()

#Filter out growth w < 100 base
covid_country_ets$conf_growth_rate_ma[covid_country_ets$confirmed <100] <- NA
covid_country_ets$death_growth_rate_ma[covid_country_ets$deaths <100] <- NA
covid_country_ets$recovered_growth_rate_ma[covid_country_ets$recovered <100] <- NA
covid_country_ets$death_rate[covid_country_ets$deaths <100] <- NA
covid_country_ets$recovered_rate[covid_country_ets$recovered <100] <- NA

    

#Only get countries with >1000 cases today
current_cases <- covid_country_ets %>% group_by(Country.Region) %>% summarise(current_cases = max(confirmed, na.rm=TRUE))
five_hundred <- current_cases$Country.Region[current_cases$current_cases >= 1000]

covid_country_ets <- covid_country_ets %>% filter(Country.Region %in% five_hundred)

#Create a panel data set
library(plm)
covid_plm <- pdata.frame(covid_country_ets, index=c("Country.Region", "date"))
#create diff variables
covid_plm$diff_confirmed <- diff(covid_plm$confirmed, 1)
covid_plm$diff_deaths <- diff(covid_plm$deaths, 1)
covid_plm$diff_recovered <- diff(covid_plm$recovered, 1)

write.csv(covid_plm, "covid_plm_country.csv")

#######
```
# Time Series by Country

## Confirmed Cases
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=confirmed, group = Country.Region)) +
  geom_line() +
  facet_wrap(~ Country.Region) +
  ggtitle(paste0("Confirmed COVID19 Cases Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

## Deaths
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=deaths, group = Country.Region)) +
  geom_line() +
  facet_wrap(~ Country.Region) +
  ggtitle(paste0("Cumulative COVID19 Deaths Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

## Recovered
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=recovered, group = Country.Region)) +
  geom_line() +
  facet_wrap(~ Country.Region) +
  ggtitle(paste0("Cumulative COVID19 Recoveries Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

## Confirmed Growth Rate 5-Day Moving Average
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=conf_growth_rate_ma, group = Country.Region)) +
  geom_line() +
  facet_wrap(~ Country.Region) +
  ggtitle(paste0("5-Day MA Confirmed Growth Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

## Death Rate 
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=death_rate, group = Country.Region)) +
  geom_line() +
  facet_wrap(~ Country.Region) +
  ggtitle(paste0("Death Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

## Recovery Rate
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm, aes(x=date, y=recovered_rate, group = Country.Region)) +
  geom_line() +
  facet_wrap(~ Country.Region) +
  ggtitle(paste0("Recovery Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```


```{r echo = FALSE, message=FALSE, warning=FALSE}
covid_ets_start <- data.frame(covid_ets)
covid_ets_state <- covid_ets_start %>% filter(Country.Region == "US") 

covid_ets_state <- covid_ets_state  %>% 
  group_by(Province.State, date) %>% 
  summarise(confirmed = sum(confirmed, na.rm=TRUE), deaths = sum(deaths, na.rm=TRUE), recovered= sum(recovered, na.rm=TRUE)) %>% 
  mutate(
    #create daily percent growth variables
    confirmed_growth_rate = c(NA,diff(confirmed))/lag(confirmed, 1),
    death_growth_rate = c(NA,diff(deaths))/lag(deaths, 1),
    recovered_growth_rate = c(NA,diff(recovered))/lag(recovered, 1),
    death_rate = deaths / confirmed,
    recovered_rate = recovered / confirmed
   
  )

#replace NaNs w 0s
covid_ets_state$confirmed_growth_rate[is.nan(covid_ets_state$confirmed_growth_rate)] <- 0
covid_ets_state$death_growth_rate[is.nan(covid_ets_state$death_growth_rate)] <- 0
covid_ets_state$recovered_growth_rate[is.nan(covid_ets_state$recovered_growth_rate)] <- 0

covid_ets_state <- covid_ets_state %>% mutate(
    conf_growth_rate_ma=rollapply(confirmed_growth_rate,5,mean,align='right',fill=NA),
    death_growth_rate_ma=rollapply(death_growth_rate,5,mean,align='right',fill=NA),
    recovered_growth_rate_ma=rollapply(recovered_growth_rate,5,mean,align='right',fill=NA)) %>%
ungroup()


#Filter out growth w < 25 base
covid_ets_state$conf_growth_rate_ma[covid_ets_state$confirmed <100] <- NA
covid_ets_state$death_growth_rate_ma[covid_ets_state$deaths <25] <- NA
covid_ets_state$recovered_growth_rate_ma[covid_ets_state$recovered <25] <- NA
covid_ets_state$death_rate[covid_ets_state$deaths <25] <- NA
covid_ets_state$recovered_rate[covid_ets_state$recovered <25] <- NA


#Only get states with >50 cases today
current_cases <- covid_ets_state %>% group_by(Province.State) %>% summarise(current_cases = max(confirmed, na.rm=TRUE))
hundred <- current_cases$Province.State[current_cases$current_cases >= 100]

covid_ets_state <- covid_ets_state %>% filter(Province.State %in% hundred)

#Create a panel data set
library(plm)
covid_plm_state <- pdata.frame(covid_ets_state, index=c("Province.State", "date"))
#create diff variables
covid_plm_state$diff_confirmed <- diff(covid_plm_state$confirmed, 1)
covid_plm_state$diff_deaths <- diff(covid_plm_state$deaths, 1)
covid_plm_state$diff_recovered <- diff(covid_plm_state$recovered, 1)

write.csv(covid_plm_state, "covid_plm_state.csv")

```



# Time Series by State

We only look at states with more than one hundred cases today. For moving average growth rates, we only look at states with deaths and recoveries over 25.

## Confirmed Cases
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm_state, aes(x=date, y=confirmed, group = Province.State)) +
  geom_line() +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Confirmed COVID19 Cases Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

## Deaths
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm_state, aes(x=date, y=deaths, group = Province.State)) +
  geom_line() +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Cumulative COVID19 Deaths Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

## Recovered
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm_state, aes(x=date, y=recovered, group = Province.State)) +
  geom_line() +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Cumulative COVID19 Recoveries Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank())
```

## Confirmed Growth Rate 5-Day Moving Average
```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm_state, aes(x=date, y=conf_growth_rate_ma, group = Province.State)) +
  geom_line() +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("5-Day MA Confirmed Growth Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

## Death Rate 5-Day Moving Average

Only states with >25 deaths are shown

```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm_state, aes(x=date, y=death_rate, group = Province.State)) +
  geom_line() +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Death Rate Through " , format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

## Recoveries Growth Rate 5-Day Moving Average

Only states with 25+ recoveries are shown

```{r echo = FALSE, message=FALSE, warning=FALSE}
ggplot(covid_plm_state, aes(x=date, y=recovered_rate, group = Province.State)) +
  geom_line() +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Recovery Rate Through ", format(Sys.time(), '%d %B %Y'))) +
  theme(axis.text.x = element_blank()) +
  theme(axis.ticks.x=element_blank()) +
  theme(axis.title.x = element_blank()) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

# Log / Time for States After 50th Case

```{r echo = FALSE, message=FALSE, warning=FALSE}
covid_plm_state_log <- covid_plm_state %>%
  filter(confirmed >=50) %>% 
  mutate(confirmed_log = log10(confirmed), deaths_log = log10(deaths), recovered_log = log10(recovered)) %>%
  group_by(Province.State) %>% 
  mutate(days_since_50 = dplyr::row_number())

```

## Log-10 by States: Confirmed Cases by Day After 50th Confirmed Case

```{r echo = FALSE, message=FALSE, warning=FALSE}
require(scales)
ggplot(covid_plm_state_log, aes(x=days_since_50, y=confirmed_log, group = Province.State, color = Province.State)) +
  geom_line(show.legend = FALSE) +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Log-10 of Confirmed Cases Since 50th Case by State as of ", format(Sys.time(), '%d %B %Y'))) +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
  labels = trans_format("log10", math_format(10^.x)))
```

## Log-10 by States: Deaths by Day After 50th Confirmed Case

```{r echo = FALSE, message=FALSE, warning=FALSE}
covid_plm_state_log_deaths <- covid_plm_state %>%
  filter(deaths >=50) %>% 
  mutate(confirmed_log = log10(confirmed), deaths_log = log10(deaths), recovered_log = log10(recovered)) %>%
  group_by(Province.State) %>% 
  mutate(days_since_50 = dplyr::row_number())

```

```{r echo = FALSE, message=FALSE, warning=FALSE}
require(scales)
ggplot(covid_plm_state_log_deaths, aes(x=days_since_50, y=deaths_log, group = Province.State, color = Province.State)) +
  geom_line(show.legend = FALSE) +
  facet_wrap(~ Province.State) +
  ggtitle(paste0("Log-10 of Deaths Since 50th Case by State as of ", format(Sys.time(), '%d %B %Y'))) +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
              labels = trans_format("log10", math_format(10^.x)))
```